Kindly Bent to Free Us

### 2020-06-02 kindly/lang.tex
Small fix.
Add some explanation.
Address a few more comments.

In the surface language (\cref{motivation}) region annotations are optional.
In the internal language, regions must be syntactically explicit and
annotated with a nesting index and a scoped variable.
This section defines a transformation $\RannotT{p}{p'}$ which
automatically inserts region annotations in  programs.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$ such that no borrow occurs
outside a region.
We give an informal presentation of our code transformation and defer
the complete definition to \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region must be contained in the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item The variable $x$ cannot occur in the region of $\borrow{x}$.
\end{enumerate}
The transformation starts from each borrow $\borrow x$ and grows its associated
region until enlarging it would include the binding for $x$ or lead to a
conflicting use of $x$ or a conflicting borrow for $x$.
As an example, consider the following program:
%
\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \lam{a}{}\ &
  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
  &g\ (\borrow[\kaff]{x});\\
  &f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})\\
\end{aligned}
}{
\begin{aligned}
  \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
  &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
  &\region[2]{\Sone x\IBORROW}{f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})}\\
  &\!\!\!}
\end{aligned}
}
\]
%
As variable $a$ only has one borrow, its region covers its whole lexical scope.
Variable $x$ has multiple conflicting borrows and requires more consideration.
We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal: making any region bigger would cause a borrowing
error.
Region indices are assigned after placement, so it is trivial to
ensure well-nested regions where the inner regions have
higher indices than the outer ones.

@@

To avoid distraction, this section focuses on the essential and novel
parts of the type system. A complete
description is available in \cref{appendix:sdtyping}.
% In this section, we focus on chosen parts of our system
% through a set of selected rules.
%
Here we only discuss the following judgments:
\begin{flushleft}
  \begin{tabular}{l@{ --- }l}
    $\inferS{C}{\E}{e}{\tau}$&
                               Expression $e$ has type $\tau$ in
                               environment $\E$ under constraints
                        $C$. \\
    $\inferSK{C}{\E}{\tau}{k}$&
                                Type $\tau$ has kind $k$ in
                                environment $\E$ under constraints
                                $C$. \\
    $\entail{D}{C}$ & Constraint $D$ entails constraint $C$.
  \end{tabular}
\end{flushleft}

@@

\begin{figure*}[tp]
  % \begin{minipage}{0.7\linewidth}
    \begin{mathpar}
      \ruleInstance
      \and
      \ruleSDVar
      \and
      \ruleSDLam
      \and
      \ruleSDApp
      \and
      \ruleSDRegion
      \and
      \ruleSDBorrow
      \and
      \ruleSDReBorrow
      \and
      \inferrule[BorrowBinding]{
        \entail{C}{(\BORROW_n\lk k) \wedge (k \lk \BORROW_\infty)}\\
        \BORROW\in\left\{\kun,\kaff\right\}
      }{
        \lregion{C}{}
        {\svar[\BORROW]{x}{\schm}^n}
        {\bbvar x k \schm}
        % {\bvar{\borrow[\BORROW]{x}}{\borrowty[\BORROW] k{\schm}}}
      }
      % \and
      % \ruleSDPair
      % \and
      % \ruleSDMatchPair
    \end{mathpar}
    \vspace{-10pt}
    \caption{Selected typing rules ($\inferS{C}{\E}{e}{\tau}$)
      and borrowing rules ($\lregion{C}{x}{\E}{\E'}$)}
    \label{selectrules:borrow}
    \label{selectrules:binders}
    \label{sdtyp:app}
    \label{selectrules:region}
    \label{env:rule:borrow}
    \vspace{-5pt}
  % \end{minipage}
\end{figure*}

>>>>

In the surface language (\cref{motivation}) region annotations are optional.
In the internal language, regions must be syntactically explicit and
annotated with a nesting index and a scoped variable.
This section defines a transformation $\RannotT{p}{p'}$ which
automatically inserts region annotations in  programs.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$ such that no borrow occurs
outside a region.
We give an informal presentation of our code transformation and defer
the complete definition to \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region must be contained in the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item The variable $x$ cannot occur in the region of $\borrow{x}$.
\end{enumerate}
The transformation starts from each borrow $\borrow x$ and grows its associated
region until enlarging it would include the binding for $x$ or lead to a
conflicting use of $x$ or a conflicting borrow for $x$.
As an example, consider the following program:
%
\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \lam{a}{}\ &
  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
  &g\ (\borrow[\kaff]{x});\\
  &f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})\\
\end{aligned}
}{
\begin{aligned}
  \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
  &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
  &\region[2]{\Sone x\IBORROW}{f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})}\\
  &\!\!\!}
\end{aligned}
}
\]
%
As variable $a$ only has one borrow, its region covers its whole lexical scope.
Variable $x$ has multiple conflicting borrows and requires more consideration.
We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal: making any region bigger would cause a borrowing
error.
In particular, it is essential to place the second borrow around \emph{both}
occurrence of $(\borrow[\kun]{x}$: we want the function to receive both
borrows, but its result must not contain any borrow (otherwise it would escape).
Region indices are assigned after placement, so it is trivial to
ensure well-nested regions where the inner regions have
higher indices than the outer ones.

@@

To avoid distraction, this section focuses on the essential and novel
parts of the type system. A complete
description is available in \cref{appendix:sdtyping}.
% In this section, we focus on chosen parts of our system
% through a set of selected rules.
%
Here we only discuss the following judgments:
\begin{flushleft}
  \begin{tabular}{l@{ --- }l}
    $\inferS{C}{\E}{e}{\tau}$&
                               Expression $e$ has type $\tau$ in
                               environment $\E$ under constraints
                        $C$. \\
    $\inferSK{C}{\E}{\tau}{k}$&
                                Type $\tau$ has kind $k$ in
                                environment $\E$ under constraints
                                $C$. \\
    $\entail{D}{C}$ & Constraint $D$ entails constraint $C$.\\
    $D \equivC C$ & Constraints $C$ and $D$ are equivalent.
  \end{tabular}
\end{flushleft}

@@

\begin{figure*}[tp]
  % \begin{minipage}{0.7\linewidth}
    \begin{mathpar}
      \ruleInstance
      \and
      \ruleSDVar
      \and
      \ruleSDLam
      \and
      \ruleSDApp
      \and
      \ruleSDRegion
      \and
      \ruleSDBorrow
      \and
      \ruleSDReBorrow
      \and
      \inferrule[BorrowBinding]{
        \entail{C}{(\BORROW_n\lk k) \wedge (k \lk \BORROW_\infty)}\\
        \BORROW\in\left\{\kun,\kaff\right\}
      }{
        \lregion{C}{x}
        {\svar[\BORROW]{x}{\schm}^n}
        {\bbvar x k \schm}
        % {\bvar{\borrow[\BORROW]{x}}{\borrowty[\BORROW] k{\schm}}}
      }
      % \and
      % \ruleSDPair
      % \and
      % \ruleSDMatchPair
    \end{mathpar}
    \vspace{-10pt}
    \caption{Selected typing rules ($\inferS{C}{\E}{e}{\tau}$)
      and borrowing rules ($\lregion{C}{x}{\E}{\E'}$)}
    \label{selectrules:borrow}
    \label{selectrules:binders}
    \label{sdtyp:app}
    \label{selectrules:region}
    \label{env:rule:borrow}
    \vspace{-5pt}
  % \end{minipage}
\end{figure*}
